Conditional resampling improves calibration \\ in single cell CRISPR screen analysis

### 2020-12-03 sceptre_paper/manuscript/manuscript_v3.tex
remove reliance on defunct ravel package.

We reanalyzed the Gasperini et al.\ data using the improved negative binomial regression approach. In addition to sequencing depth, we included as covariates in the regression model the total number of expressed genes per cell, as well as the technical factors that Gasperini et al.\ accounted for in their analysis (namely, total number of gRNAs detected per cell, percentage of transcripts mapped to mitochondrial genes, and sequencing batch). We also applied the improved negative binomial approach to the Xie et al.\ ARL15-enh data. We included sequencing depth, number of detected gRNA UMIs per cell, and sequencing batch as covariates in the analysis of the latter dataset. The improved negative binomial method exhibited better calibration than the Monocle negative binomial method on the Gasperini data (Figure \ref{fig:ground_truth}b). However, on both the Gasperini et al.\ data and Xie et al.\ data, the improved negative binomial method showed clear signs of p-value inflation (Figure \ref{fig:ground_truth}b-c).

@@

\paragraph{SCEPTRE produces correct results when the ground truth is known.}

We assessed the calibration, sensitivity, and specificity of SCEPTRE in several settings in which the ground truth was known. First, we demonstrated the calibration of SCEPTRE in a small, proof-of-concept simulation study (Figure \ref{fig:ground_truth}a). We considered a class of negative binomial regression models with dispersion $\alpha$ and two technical covariates (sequencing depth and batch). We simulated expression data for a single gene in 1000 cells using four models selected from this class: the first with $\alpha = 1$, the second with $\alpha = 0.25$, the third with $\alpha = 5$, and the last with $\alpha = 1,$ but with 25$\%$ zero-inflation. We also simulated negative control gRNA data using a logistic regression model with the same covariates as the gene expression model. We assessed the calibration of three methods across the four simulated datasets: SCEPTRE, improved negative binomial regression, and scMAGeCK-LR,\cite{Yang2020} a recently-proposed, permutation-based nonparametric method. To assess the impact of model misspecification on SCEPTRE and the improved negative binomial method (on which SCEPTRE relies), we fixed the dispersion of the negative binomial method to 1 across all four simulated datasets. We found that the negative binomial method works as expected when the model is correctly specified but breaks down in all three cases of model misspecification. scMAGeCK-LR exhibits poor calibration across all simulated datasets, likely because, as a traditional nonparametric method, it fails to adequately account for confounders. Finally, SCEPTRE is well-calibrated regardless of expression model misspecification and confounder presence.

@@

\clearpage
\begin{figure*}[h!]
	\includegraphics[width = \textwidth]{figures/Figure3/Figure3_cropped.pdf}
 \end{figure*}
\captionof{figure}[short caption here]{
	\label{fig:ground_truth}
	\textbf{SCEPTRE demonstrates excellent calibration, specificity, and sensitivity under known ground truth.} \textbf{a,} Numerical simulation comparing three methods -- SCEPTRE, improved negative binomial regression, and scMAGeCK-LR -- on four simulated datasets. Only SCEPTRE maintains calibration despite model misspecification and confounder presence. \textbf{b-c},  	Application of SCEPTRE, improved negative binomial regression, Monocle regression, and Virtual FACS to (b) all negative control gRNAs and all genes in the Gasperini et al.\ data, and (c) all ARL15-enh-targeting gRNAs and all genes in the Xie et al.\ data. Compared to the other methods, SCEPTRE shows excellent calibration. Gene \textit{ARL15} annotated in (c). \textbf{d,} SCEPTRE $p$-values for Gasperini et al.\ TSS-targeting controls are highly significant, and in general, more significant than the original empirical $p$-values. \textbf{e,} Comparison of $p$-values produced by SCEPTRE for ARL15-enh to $p$-values produced by an arrayed, bulk RNA-seq CRISPR screen of ARL15-enh. The results of the two analyses coincide exactly, with both aanlyses rejecting gene \textit{ARL15} (and only gene \textit{ARL15}) after a Benjamini-Hochberg correction. Dotted blue lines, rejection thresholds.}
\noindent\hrulefill

@@

\section*{Data availability}

>>>>

We reanalyzed the Gasperini et al.\ data using the improved negative binomial regression approach. In addition to sequencing depth, we included as covariates in the regression model the total number of expressed genes per cell, as well as the technical factors that Gasperini et al.\ accounted for in their analysis (namely, total number of gRNAs detected per cell, percentage of transcripts mapped to mitochondrial genes, and sequencing batch). We also applied the improved negative binomial approach to the Xie et al.\ ARL15-enh data. We included sequencing depth, number of detected gRNA UMIs per cell, and sequencing batch as covariates in the analysis of the latter dataset. The improved negative binomial method exhibited better calibration than the Monocle negative binomial method on the Gasperini et al.\ data (Figure \ref{fig:ground_truth}b). However, on both the Gasperini et al.\ data and Xie et al.\ data, the improved negative binomial method showed clear signs of p-value inflation (Figure \ref{fig:ground_truth}b-c).

@@

\paragraph{SCEPTRE shows good calibration, sensitivity, and specificity on real and simulated data.}

%We assessed the calibration, sensitivity, and specificity of SCEPTRE in several settings in which the ground truth was known.
First, we demonstrated the calibration of SCEPTRE in a small, proof-of-concept simulation study (Figure \ref{fig:ground_truth}a). We considered a class of negative binomial regression models with dispersion $\alpha$ and two technical covariates (sequencing depth and batch). We simulated expression data for a single gene in 1000 cells using four models selected from this class: the first with $\alpha = 1$, the second with $\alpha = 0.2$, the third with $\alpha = 5$, and the last with $\alpha = 1,$ but with 25$\%$ zero-inflation. We also simulated negative control gRNA data using a logistic regression model with the same covariates as the gene expression model. We assessed the calibration of three methods across the four simulated datasets: SCEPTRE, improved negative binomial regression, and scMAGeCK-LR,\cite{Yang2020} a recently-proposed, permutation-based nonparametric method. To assess the impact of model misspecification on SCEPTRE and the improved negative binomial method (on which SCEPTRE relies), we fixed the dispersion of the negative binomial method to 1 across all four simulated datasets. We found that the negative binomial method works as expected when the model is correctly specified but breaks down in all three cases of model misspecification. scMAGeCK-LR exhibits poor calibration across all simulated datasets, likely because, as a traditional nonparametric method, it fails to adequately account for confounders. Finally, SCEPTRE is well-calibrated regardless of expression model misspecification and confounder presence.

@@

\clearpage
\begin{figure*}[h!]
	\includegraphics[width = \textwidth]{figures/Figure3/Figure3_cropped.pdf}
 \end{figure*}
\captionof{figure}[short caption here]{
	\label{fig:ground_truth}
	\textbf{SCEPTRE demonstrates good calibration, specificity, and sensitivity under known ground truth.} \textbf{a,} Numerical simulation comparing three methods -- SCEPTRE, improved negative binomial regression, and scMAGeCK-LR -- on four simulated datasets. Only SCEPTRE maintains calibration despite model misspecification and confounder presence. \textbf{b-c},  	Application of SCEPTRE, improved negative binomial regression, Monocle regression, and Virtual FACS to (b) all negative control gRNAs and all genes in the Gasperini et al.\ data, and (c) all ARL15-enh-targeting gRNAs and all genes in the Xie et al.\ data. Compared to the other methods, SCEPTRE shows excellent calibration. Gene \textit{ARL15} annotated in (c). \textbf{d,} SCEPTRE $p$-values for Gasperini et al.\ TSS-targeting controls are highly significant, and in general, more significant than the original empirical $p$-values. \textbf{e,} Comparison of $p$-values produced by SCEPTRE for ARL15-enh to $p$-values produced by an arrayed, bulk RNA-seq CRISPR screen of ARL15-enh. The results of the two analyses coincide exactly, with both aanlyses rejecting gene \textit{ARL15} (and only gene \textit{ARL15}) after a Benjamini-Hochberg correction. Dotted blue lines, rejection thresholds.}
\noindent\hrulefill

@@

\clearpage
\section*{Data availability}
